Annotation-Based Empirical Performance Tuning Using Orio

### 2013-11-05 doc/papers/ipdps09/integration.tex

Figure~\ref{fig:pluto-orio} outlines the overall structure of the
Pluto-Orio integrated system, which is implemented as a new
optimization module in Orio. Initially, the transformation process
takes two kinds of input: the loop code to be optimized and the
optimization parameters code. The loop code is embedded in the
annotation body block, while the performance parameters code is
written in the module body block. A simple parser implemented inside
the new module parses the performance parameters code to extract their
values. The extracted values of the performance parameters include
tile sizes, unroll factors, loop order (permutation), as well as
several boolean values for triggering OpenMP parallelization, scalar
replacement, and auto-vectorization. Using these parameter values,
Pluto then performs polyhedral transformations (e.g., two-level tiling
and OpenMP parallelization) on the input loop code. The resulting
generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. The
statistical information produced by gprof provides accurate locations
(line numbers) of the loop nests where the Pluto-generated code spends
most of its execution time. The identified hotspot loop nests are then
decorated with Orio's performance annotations for complementary
syntactic transformations (e.g., loop permutation, loop
unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as
described in Section~\ref{sec:implm}.

Pluto also performs syntactic loop unroll/jamming as a post-processing
pass. The target loops, however, are limited only to innermost loops with a
maximum depth of two (i.e., 1-D unrolling and 2-D unroll/jamming). So we
choose to use the loop unroll/jamming transformation already available in
Orio, which is more flexible because it can be applied to loop nests of
depths larger than two.

At present Orio does not employ any data dependency analysis when performing
syntactic transformations. Therefore, there is no guarantee that the code
generated after syntactic transformations is correct. Because of this, the
tuning process using the integrated Pluto-Orio system is currently
semi-automatic---user involvement is required to decide whether it is safe
to apply a syntactic transformation on an identified hotspot.
%The user only needs
%to check the transformation legalities once, since the hotspots
%consistently have similar basic loop structure for different
%Pluto-generated code versions. Once all illegal syntactic
%transformations are known by the user, they must be switched off so
%that the tuning process can proceed fully automatic with no more human
%interventions.

>>>>

Figure~\ref{fig:pluto-orio} outlines the overall structure of the
Pluto-Orio integrated system, which is implemented as a new
optimization module in Orio. Initially, the transformation process
takes two kinds of input: the loop code to be optimized and the
optimization parameters code. The loop code is embedded in the
annotation body block, while the performance parameters code is
written in the module body block. A parser implemented inside
the new module parses the performance parameters code to extract their
values. The extracted values of the performance parameters include
tile sizes, unroll factors, loop order (permutation), as well as
several boolean values for triggering OpenMP parallelization, scalar
replacement, and auto-vectorization. Using these parameter values,
Pluto then performs polyhedral transformations (e.g., two-level tiling
and OpenMP parallelization) on the input loop code. The resulting
generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. The
statistical information produced by gprof provides accurate locations
(line numbers) of the loop nests where the Pluto-generated code spends
most of its execution time. The identified hotspot loop nests are then
decorated with Orio's performance annotations for complementary
syntactic transformations (e.g., loop permutation, loop
unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as
described in Section~\ref{sec:implm}.

Pluto also performs syntactic loop unroll/jam in a post-processing
pass. The target loops, however, are limited only to innermost loops with a
maximum depth of two (i.e., 1-D unrolling and 2-D unroll/jamming). So we
choose to use the loop unroll/jam transformation already available in
Orio, which is more flexible because it can be applied to loop nests of
depths larger than two.

At present Orio does not employ any data dependence analysis when performing
syntactic transformations. Therefore, there is no guarantee that the code
generated after syntactic transformations is correct. Because of this, the
tuning process using the integrated Pluto-Orio system is currently
semi-automatic---user involvement is required to decide whether it is safe
to apply a syntactic transformation on an identified hotspot.
%The user only needs
%to check the transformation legalities once, since the hotspots
%consistently have similar basic loop structure for different
%Pluto-generated code versions. Once all illegal syntactic
%transformations are known by the user, they must be switched off so
%that the tuning process can proceed fully automatic with no more human
%interventions.

### 2013-11-05 doc/papers/ipdps09/related.tex

An alternative to manual or automated tuning of application codes is
the use of tuned libraries. The two basic approaches to supplying
high-performance libraries include providing a library of hand-coded
options (e.g., \cite{BLAS,ESSL,Goto:2006fk}) and generating optimized
code automatically for the given problem and machine
parameters. ATLAS~\cite{atlas_sc98,WN147} for BLAS~\cite{BLAS} and
some LAPACK~\cite{laug} routines, OSKI~\cite{OSKI} for sparse linear
algebra, PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products,
and domain-specific libraries such as FFTW~\cite{frigo98} and
SPIRAL~\cite{SPIRAL} are all examples of the latter approach. Most
automatic tuning approaches perform empirical parameter searches on
the target platform.  FFTW uses a combination of static models and
empirical techniques to optimize FFTs. SPIRAL generates optimized
digital signal processing libraries by an extensive empirical search
over implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on
the other hand, achieves great performance results on several
architectures by using hand-tuned data structures and kernel
operations.  These auto- or hand-tuned approaches can deliver
performance that can be five times as fast as that produced by many
optimizing compilers
\cite{WN147}.  The library approach, however, is limited by the fact
that optimizations are highly problem- and
machine-dependent. Furthermore, at this time, the functionality of the
currently available automated tuning systems is quite limited.

@@

Emerging annotation-based tools are normally designed by compiler researchers
and thus the interfaces are not necessarily based on concepts accessible to
computational scientists. The complexity of existing annotation languages and lack
of common syntaxes for transformations (e.g., loop unrolling) result
in steep learning curves and the inability to take advantage of more than one
approach at a time. Furthermore, at present, there is no good way for
users to learn about the tools available and compare their
capabilities and performance.

>>>>

An alternative to manual or automated tuning of application codes is
the use of tuned libraries. The two basic approaches to supplying
high-performance libraries include providing a library of hand-coded
options (e.g., \cite{BLAS,ESSL,Goto:2006fk}) and generating optimized
code automatically for the given problem and machine
parameters. ATLAS~\cite{atlas_sc98,WN147} for BLAS~\cite{BLAS} and
some LAPACK~\cite{laug} routines, OSKI~\cite{OSKI} for sparse linear
algebra, PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products,
and domain-specific libraries such as FFTW~\cite{frigo98} and
SPIRAL~\cite{SPIRAL} are all examples of the latter approach. Most
automatic tuning approaches perform empirical parameter searches on
the target platform.  FFTW uses a combination of static models and
empirical techniques to optimize FFTs. SPIRAL generates optimized
digital signal processing libraries by an extensive empirical search
over implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on
the other hand, achieves near-peak performance on several
architectures by using hand-tuned data structures and kernel
operations.  These auto- or hand-tuned approaches can deliver
performance that can be five times as fast as that produced by many
optimizing compilers
\cite{WN147}.  The library approach, however, is limited by the fact
that optimizations are highly problem- and
machine-dependent. Furthermore, at this time, the functionality of the
currently available automated tuning systems is quite limited.

@@

Emerging annotation-based tools are normally designed by compiler researchers
and thus the interfaces are not necessarily based on concepts accessible to
computational scientists. The complexity of existing annotation languages and lack
of common syntax for transformations (e.g., loop unrolling) result
in steep learning curves and the inability to take advantage of more than one
approach at a time. Furthermore, at present, there is no good way for
users to learn about the tools available and compare their
capabilities and performance.
